---
title: "Final Exam"
format: 
  html:
    self-contained: true
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

Please read all of the questions and tables carefully and follow all instructions. Each question has an allotted point value. Be as thorough and specific as possible; extra calculations and incorrect information, even in the presence of correct information, will result in point deductions. Be sure to show all formulas and all necessary work to answer the questions. You may upload your completed exam to the Elms course site.

::: callout-note
While this is an open-note exam, you are not to receive assistance from anyone else. As usual, the Honor Code applies.
:::

**Total points: 50**

## Conceptual Questions

Please include all work (and computations) necessary to answer the questions.

**Total points: 20 (2 points each)**

### Question 1

Suppose you estimate a bivariate regression model with a total sample size of 50 and obtain a standard error for $\beta_1$ of 2.50. What is the smallest regression coefficient ($\beta_1$) that you could have and still be able to reject the null hypothesis (of no relationship between $X_1$ and $Y$) at the 0.05 level with a one-tailed test?

```{r}
sample_size <- 50
se_beta1 <- 2.50
alpha <- 0.05

df <- sample_size - 1 - 1

critical_value <- qt (1 - alpha, df)

smallest_beta1 <- critical_value * se_beta1

print(smallest_beta1)
```

### Question 2

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and find that $\beta_1 = 4.20$, $SE_{\beta_1} = 1.90$, $t = 2.21$, $p = 0.01$, and a 95% confidence interval is \[0.48, 7.92\]. What is my best estimate of the effect of a two-unit change in $X$ on the mean of $Y$ in the population?

Effect on Y = beta_1 \* 2 = 4.20 \* 2 = 8.40

### Question 3

I conduct an OLS regression with a sample size of 90 and 5 independent variables. To determine a p-value for each coefficient, I would examine a t-distribution with how many degrees of freedom?

Degree of freedom = sample size - number of IVs - 1

= 90 - 5 - 1 = 84

### Question 4

Suppose I regress $Y$ on $X$ and compute the mean response for $Y$ at some specified value of $X$. When determining the confidence interval around this mean response, which of the following will **NOT** have any effect on the width (or, size) of that confidence interval?

A.  Total sample size,

B.  Mean-squared error,

C.  The specified value of $X$,

D.  None of the above: all of these (above) will affect the confidence interval,

E.  There is not enough information to answer this question.

**Answer: C**

### Question 5

I regress $Y$ on $X$ and find that $\beta_1$ has a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  The lower and upper bound of a 95% confidence interval around $\beta_1$ will have the same sign,

B.  The absolute value for the t-statistic for $\beta_1$ will be greater than 1.96,

C.  A 90% confidence interval around $\beta_1$ will not contain zero,

D.  All of the above,

E.  None of the above and/or there is not enough information.

**Answer: D**

### Question 6

I regress $Y$ on three independent variables -- $X_1$, $X_2$, and $X_3$ -- and I find the following 95% confidence intervals -- $\beta_1$: \[0.12, 1.45\], $\beta_2$: \[-0.01, 0.15\], and $\beta_3$: \[-0.64, -0.01\]. Which of the following statements is most accurate?

A.  $\beta_2$ and $\beta_3$ will have negative coefficients, and $\beta_1$ is statistically significant at the 0.05 level (two-tailed),

B.  $\beta_1$ will have a positive coefficient, and $\beta_1$ is the only statistically significant coefficient (of the three coefficients) at the 0.05 level (two-tailed),

C.  $\beta_1$ will have a positive coefficient, $\beta_2$ and $\beta_3$ will have negative coefficients, and only $\beta_1$ and $\beta_3$ are statistically significant at the 0.05 level (two-tailed),

D.  All of the above,

E.  None of the above and/or there is not enough information.

**Answer: C**

### Question 7

Suppose I estimate a regression with two independent variables and obtain a $R^2$ of 0.40 where the Residual Sum of Squares is equal to 5,150. What does the Total Sum of Squares equal in this regression model?

```{r}
r_squared <- 0.40  
RSS <- 5150

TSS <- RSS / (1 - r_squared)

print(TSS)
```

### Question 8

I regress $Y$ on $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$). I find that $\beta_1 = -0.87$ with a two-tailed p value of 0.001 and $\beta_2 = 3.52$ with a two-tailed p-value of 0.04. Which of the following statements is most accurate?

A.  $X_2$ has a larger substantive effect on $Y$ than $X_1$, but the effect of $X_1$ is more statistically significant,

B.  $X_1$ has a larger substantive effect on $Y$ than $X_2$ and $X_1$ is more statistically significant than $X_2$,

C.  $X_1$ has a larger substantive effect on $Y$ than $X_2$, but the effect of $X_2$ is more statistically significant,

D.  None of the above and/or there is not enough information.

**Answer: A**

### Question 9

I regress $Y$ on $X_1$ (i.e., $Y = \beta_0 + \beta_1 X_1$) and obtain a $R^2$ of 0.45. Then, I regress $Y$ on both $X_1$ and $X_2$ (i.e., $Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2$) and obtain a $R^2$ of 0.53. Which of the following statements must be true?

A.  The correlation between $X_1$ and $Y$ is stronger than the correlation between $X_2$ and $Y$,

B.  The coefficient on $\beta_1$ is statistically significant in both models,

C.  The model with $X_1$ and $X_2$ explains more of the variation in $Y$ than a model with just $X_1$

D.  All of the above: all of these statements are true.

**Answer: C**

### Question 10

Based on the following regression equation, might the negative coefficient on the South variable be the result of southern states having a lower proportion of high school graduates than non-southern states? In no more than one sentence, explain your answer.

$$
Turnout = 34 + 0.5(Percent\ High\ School\ Graduates) – 5.9(South)
$$

No, the negative coefficient on the South variable does not necessarily imply causation/effect from the proportion of high school graduates, as the regression only indicates an association between being in the South and lower turnout, controlling for the percentage of high school graduates.

## Applied Questions

All datasets referenced below are posted on the ELMs course site.

### Question 1

*Points: 15*

Use the `world` dataset to answer the following questions. The relevant variables and their coding information are as follows: `literacy` indicates a country's literacy rate; `dem_score14` represents a country's level of democratization (higher values indicate greater democratization); `spendeduc` reflects the amount of public expenditures on education as a percentage of GDP; `gdp_10_thou` represents GDP per capita (one unit is \$10,000 USD); `educ_quality` indicates the average quality rating of a country's educational system; and `ungr9095` represents the percent average annual population growth.

```{r}
library(stargazer)
library(car)
library(dplyr)
library(ggplot2)
library(interactions)
```

#### Part A

*Points: 4*

Evaluate the bivariate relationship (using R) between a country's support for public education and its literacy rate. Report the results in a professionally formatted table. Do countries that spend more money on education appear to have a higher literacy rate?

```{r}
world <- read.csv("/Users/noory211/Desktop/GVPT/Fall 2023/622 Quant Method/gvpt_problems_set/world.csv")

literacy_level <- lm (literacy ~ spendeduc, data = world)

summary (literacy_level)
```

```{r}
stargazer (literacy_level, type = "text", title = "Regression Results",
           covariate.labels = c ("Education Spending"))
```

The bivariate OLS analysis indicates that higher spending on education is associated with an increased literacy rate. For every one-unit increase in education spending, the literacy rate increases by 1.722 points, on average. This effect is significant at the 10% level, albeit relatively weak. This is confirmed by the low R-squared value (2.7%), suggesting that education spending may not be a strong predictor in explaining variations in literacy rates. The intercept values indicates that in the context of zero education spending, the literacy rate stands at 73.3% (significant at 1%).

#### Part B

*Points: 5*

Estimate a second regression model (using R) that includes both a country's public expenditures on education and its level of democratization as independent variables. When controlling for democratization, does education expenditures exhibit a significant impact on literacy? Report the results in a professionally formatted table. Be sure to discuss the extent to which the results change, and if so, why they changed.

```{r}
literacy_level <- lm (literacy ~ spendeduc + dem_score14, data = world)

summary (literacy_level)
```

```{r}
stargazer (literacy_level, type = "text", title = "Second Regression Results",
           covariate.labels = c ("Education Spending", "Level of Democratization"))
```

When we include a control variable in the model, the effect of education spending on the literacy rate disappears. However, the control variable, the country's level of democratization, is statistically significant at the 1% level. A one-unit increase in the level of democratization is associated with a 4.650-unit increase in the literacy rate, holding all other constant. The R-squared value is significantly boosted compared to the first model, suggesting that this model is a better predictor of literacy rates. This implies that countries become more literate due to democratization, rather than increased spending on education. It is highly likely that the magnitude of education spending is mediated through whether a country is democratized. More democracy leads to more investment in education, not the other way around. Therefore, the first model suffers from omitted variable bias, and it is crucial to include this control variable in the model for a better understanding of the factors influencing literacy rates.

#### Part C

*Points: 6*

Now estimate a third model (using R) that includes all of the variables listed above. Report the regression results in a professionally formatted table and interpret each coefficient. When interpreting the impact of democratization using this regression model, present/utilize a visual representation (using R) of the expected change in literacy as a function of democratization.

```{r}
literacy_level <- lm (literacy ~ spendeduc + dem_score14 + gdp_10_thou + educ_quality + ungr9095, data = world )

summary (literacy_level)
```

```{r}
stargazer (literacy_level, type = "text", title = "Third Regression Results",
           covariate.labels = c ("Education Spending", "Level of Democratization", "GDP Per Capita", "Quality of Education", "Annual Population Growth"))
```

```{r}
library(marginaleffects)
plot_predictions (literacy_level, condition = "dem_score14")

```

The final model presents a comprehensive multiple regression analysis explaining the level of literacy rates. Among the five predictors examined, at least three variables are identified as significant predictors of literacy (significant at 1% and 10% levels). The country's level of democratization and the quality of education are both positively correlated with literacy. A one-unit increase in the level of education is associated with an average increase of 2.546 points in literacy at the 1% significance level, holding all other variables constant. Similarly, a one-point increase in the quality of education is associated with an average increase of 3.812 points in literacy, holding all other variables constant. Annual population growth is the only variable that negatively impacts literacy, suggesting that for every one-unit increase in population growth, the literacy level drops by 3.326 points on average (1% level). Education spending and a country's GDP per capita do not appear to have a significant impact on literacy. The R-squared value is 36.7%, indicating a modest relationship among our explanatory variables in explaining variations in literacy levels.

We are particularly interested in exploring the relationship between a country's level of democratization and its literacy rate. As can be seen from the graph, there is a positive relationship between democratization and literacy rates. This correlation suggests that higher levels of democratization within a country may correspond to stronger literacy rates among its population.

### Question 2

*Points: 15*

```{r}
states <- read.csv("/Users/noory211/Desktop/GVPT/Fall 2023/622 Quant Method/gvpt_problems_set/states.csv")

library(modelsummary)
```

Using the `states` dataset, use R to regress the variable `obama08` (the percentage of a state's vote that President Obama received in the 2008 U.S. presidential election) on the following independent variables: `cig_tax` represents the amount of a state's cigarette tax (in dollars); `college` is the percentage of a state's population that graduated college (0 to 100); `union07` is the percentage of a state's workers that are union members (0 to 100); and `south` is a dichotomous variable coded as `1` if the state is located in the South (`0` otherwise). Report the results in a professionally formatted table. Answer the following questions.

```{r}
obama_vote <- lm (obama08 ~ cig_tax + college + union07 + south, data = states)

summary (obama_vote)
```

```{r}
modelsummary(obama_vote, stars = c ("***" = 0.01, "**" = 0.05, "*" = 0.1), coef_rename = c("cig_tax" = "Cigarette Tax", "college" = "College Graduate", "union07" = "Union Members", "south" = "Southern States"),
             statistic = c("SE = {std.error}", "conf.int"))

```

#### Part A

*Points: 6*

Interpret the substantive results from this regression model. Be sure to discuss the direction, magnitude, and statistical significance of each slope coefficient. Does the `union07` variably exhibit a substantively significant impact on voting behavior in the 2008 election?

The analysis reveals a statistically significant impact of the 'union' variable on the votes garnered by Obama in the 2008 U.S. presidential election. The positive coefficient associated with this variable suggests that an increase in the proportion of a state's workforce that is unionized is correlated with a higher likelihood of voting for Obama. A one percent increase in union membership within a state is linked to an average increase of 75 percent in votes for Obama, significant at the 1% level, while controlling for other variables.

Furthermore, the level of education, particularly the proportion of the population with a college degree, emerges as another significant predictor. The model indicates that for each one-point increase in the percentage of college graduates, there is an average boost of 96.5% in Obama's votes, significant at the 1% level and accounting for other variables.

While the variables representing cigarette tax rates and the designation of Southern states exhibit positive coefficients, their associations with Obama's support are not as pronounced. Lastly, the R-squared value of the model is 52%, denoting a moderately effective explanatory power of the chosen predictors in accounting for the variations in the dependent variable.

#### Part B

*Points: 2*

Interpret the intercept coefficient. What does this represent and is it meaningful to interpret this coefficient on its own? Why or why not?

In this model, the intercept is 16.12, representing the predicted percentage of votes that Obama would have received in the 2008 election in a non-Southern state if the cigarette tax was zero and there were zero percentages for college graduates and union membership. Although the intercept is statistically significant, interpreting it in isolation is not meaningful. This is because the hypothetical conditions it represents are non-existent in reality, rendering any interpretation nonsensical.

#### Part C

*Points: 2*

What is the expected mean percentage of Obama's vote total in a Southern state with a cigarette tax of \$2.00, a state population where 30% of people are college graduates, and where 15% of workers in a state are union members?

```{r}
# Regression coefficients
beta_0 <- 16.1280
beta_1 <- 0.5230
beta_2 <- 0.9646
beta_3 <- 0.7514
beta_4 <- 1.3928

# Given conditions
cigarette_tax <- 2.00
college_grad <- 0.30
union_members <- 0.15
southern_state <- 1  

# Calculating Obama expected votes
obama_expected_votes <- beta_0 + beta_1 * cigarette_tax + beta_2 * college_grad + beta_3 * union_members + beta_4 * southern_state

print(obama_expected_votes)
```

#### Part D

*Points: 3*

What is a 90% confidence interval around the coefficient for the `south` variable.

```{r}
coef_south <- 1.3928
se_south <- 2.5885

# Calculating degree of freedom
df <- 50 - 4 - 1 

# Finding the t-critical value for a 90% confidence interval
alpha <- 0.10  
t_critical <- qt(1 - alpha/2, df)

# Calculating the confidence interval
lower_bound <- coef_south - t_critical * se_south
upper_bound <- coef_south + t_critical * se_south

print(lower_bound)
print(upper_bound)
```

#### Part D

*Points: 3*

What is the null hypothesis that the F-test in this regression output is testing?

The null hypothesis in my regression posits that the independent variables do not exert a statistically significant effect on the dependent variable. In other words, all the regression coefficients in the model are equal to zero. However, since the p-values for two of my independent variables --- the 'college graduate' and 'union' variables --- are statistically significant, I can reject this null hypothesis. This suggests that the model, as a whole, is statistically significant in explaining Obama's votes share.
