---
title: "Problem Set 8"
subtitle: "Due date: 20 November"
format: 
  html:
    self-contained: true
toc: true
editor: visual
execute: 
  echo: true
---

Please upload your completed assignment to the ELMs course site (under the assignments menu). Remember to include an annotated script file for all work with R and show your math for all other problems (if applicable, or necessary). Please also upload your completed assignment to the Github repository that you have shared with us. *We should be able to run your script with no errors.*

**Total points: 30**

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(broom)
library(modelsummary)
library(ggdist)
library(MetBrewer)
```

## Question 1

*Points: 5*

::: {.callout-tip}
5/5
:::

For the following regression equation, $\hat{Y} = 8.5 + 6x + \epsilon$, the standard error for $\beta_0$ is 2.5, the standard error for $\beta_1$ is 3.5, and the sample size is 2000. Find the t-statistic, 95% confidence interval, and p-value (using a two-tailed test) for $\beta_1$.

Is $\beta_1$ statistically significant at the 0.05-level with a two-tailed test? Why or why not?

```{r}
beta_1 <-6
se_beta_1 <- 3.5
n <- 2000
alpha <- 0.05

#calculating the t-statistic for beta_1
t_stat <- beta_1/se_beta_1
t_stat

#calculating degrees of freedom
df <- n - 1 - 1

#5% threshold
t_stat_95 <- qt(0.025, df = df, lower.tail = F)

#calculating 95% confidence interval
lower_ci <- beta_1 - t_stat_95 * se_beta_1
upper_ci <- beta_1 + t_stat_95 * se_beta_1
lower_ci
upper_ci

#calculating p-value
p_value <- 2 * pt(t_stat, df = df, lower.tail = F)
p_value


```

t-statistic = 1.71

95% CI = \[- 0.86, 12.86\]

p-value (two-tailed) = 0.086

Since our p-value is greater than 0.05 level, beta_1 coefficient is not statistically significant. This means that there is not enough evidence to reject the null hypothesis that beta_1 = 0.

## Question 2

*Points: 5*

::: {.callout-tip}
5/5
:::

Suppose you estimate an OLS regression and retrieve a $R^2$ value of 0.45. If the Total Sum of Squares (TSS) from that regression equals 4,700, what is the value for the Residual Sum of Squares (RSS)?

```{r}
r_squared <- 0.45
TSS <- 4700

#Calculating RSS
RSS <- TSS * (1 - r_squared)
RSS
```

the Residual Sum of Squares (RSS) = 2585

## Question 3

*Points: 5*

::: {.callout-tip}
5/5
:::

Suppose you estimate a bivariate regression with a sample size of 102 and obtain a regression coefficient ($\beta_1$) of 5.0. What is the largest standard error that $\beta_1$ could have and still be statistically significant (i.e., reject the null hypothesis of no relationship) at the 0.05 level with a one-tailed test?

```{r}
n <- 102
beta_1 <- 5
alpha <- 0.05

#Calculating degrees of freedom
df <- 102 - 1 - 1

#Calculating t-value for one-tailed test
t_value <- qt(1 - alpha, df)

#Calculating the largest SE
se_max <- abs(beta_1) / t_value
se_max
```

Since beta_1 = 5 and t value in one-tailed 5% with 100 df is 1.66, then we can calculate the largest standard error by dividing 5 with 1.66 = 3.01.

## Question 4

*Points: 5*

::: {.callout-tip}
5/5
:::

Using the `states` dataset from the `poliscidata` package, produce a scatterplot of the variables `romney2012` and `hispanic10` (with `romney2012` as the dependent variable on the y-axis). Fit a regression line to the scatterplot. Describe the scatterplot and include a copy of it. Note any suspected outliers, if any (a visual inspection will suffice for this question).

::: callout-note
The variable `romney2012` measures the percentage of the state's vote that Mitt Romney received in the 2012 presidential election, and `hispanic10` indicates the percentage of the state's population that identified as Hispanic in 2010.
:::

```{r}
library(poliscidata)

ggplot(states, aes(x = hispanic10, y = romney2012)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = F) + 
  theme_minimal() 
```

The graph depicts a negative relationship between the number of Hispanic voters and Mitt Romney's vote share in the 2012 election, based on 50 states of the US. This suggests that in states with fewer Hispanic voters, Romney tended to receive more votes. Additionally, there are a few potential outliers where Romney received significantly fewer votes in states with less than 10 percent Hispanic population.

## Question 5

*Points: 10*

::: {.callout-tip}
10/10
:::

Estimate a bivariate regression with `romney2012` as the dependent variable and `hispanic10` as the independent variable and report the results in a professionally formatted table. In as much detail as possible, describe (and interpret) the regression results.

```{r}

romney <- lm (romney2012 ~ hispanic10, data = states)
summary (romney)
```

```{r}
modelsummary(romney,
             coef_rename = c("hispanic10" = "Percent of Hispanic"),
             statistic = c("std.error", "p.value"))
```

```{r}
modelplot(romney,
          coef_rename = c("hispanic10" = "Percent of Hispanic"))
```

The bivariate OLS analysis indicates that for every one percent increase in the Hispanic population, Romney's vote share decreases by an average of 19.6 percent. However, this relationship is not statistically significant, as the p-value is 0.182, which is higher than the conventional significance levels. This finding is further supported by the low R-squared value of 0.036, suggesting that the independent variable may not be a strong predictor of Romney's vote share. Although a negative relationship is observed in the scatterplot, the association between the two variables is weak, and there is insufficient evidence to assert that the size of the Hispanic minority significantly affects Romney's vote share.
